---
title: "Lab 05 — Sampling and Bootstrap Distributions"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

In EAES we almost never observe an entire population. Instead, we take **samples** and compute **point estimates** (like a mean) to infer **population parameters**.

This lab focuses on four core ideas:

1. **Relative error** of point estimates (accuracy)
2. **Standard error (SE)** and how it changes with sample size
3. How **replication** (many repeated samples) creates an approximate **sampling distribution**
4. (Intro) **Bootstrap** distributions: resampling from one sample to approximate uncertainty

## Coding expectations (Week 7+)

This lab uses a mix of:
- **Fill-in-the-blank** (quick checks)
- **Partial pipelines** (you finish the analysis)
- **From-scratch chunks** (you write the full solution)

Use slides + past labs + Discord as needed, but make sure your final code is **your own** and runs **top-to-bottom**.

---

# Data

## Load data (partial pipeline)

Complete the pipeline so `df` is a clean tibble. Filter for 2023, and replace fill values (-9999) with na.

```{r load_data, echo=TRUE, eval=FALSE}
# GOAL: Read the CSV from the repo and clean column names.
# TODO: finish this pipeline.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>% 
  filter(year_local == 2023) 

glimpse(df)
```

## Derive time columns 

Write code to add:
- `date` from `year_local` + `doy`
- `month` as labeled abbreviated month (Jan–Dec)
- `day_night` as "Day"/"Night" from `daytime` (0/1)

```{r derive_time, echo=TRUE, eval=FALSE}
# GOAL: Create date/month/day_night columns.
# HINT: DOY is 1-indexed, so use (doy - 1) with origin "YYYY-01-01".

df <- df %>%
 mutate(
    # TODO: create a Date column from year_local and doy
    # HINT: Jan 1 is DOY = 1, so use (doy - 1) with origin = "YYYY-01-01"
    date = as.Date(doy - 1, origin = paste0(year_local, "-01-01")),

    # TODO: create a month column (numeric 1–12 or labeled months)
    month = month(date, label = TRUE, abbr = TRUE),

    # TODO: make a day/night label using daytime (0/1)
    day_night = if_else(daytime == 1, "Day", "Night")
  )

df %>% count(month)
df %>% count(day_night)
```


# Population parameters vs point estimates

## 1) Choose variables (fill-in)

Pick **2–3 variables** to analyze. Include at least **one flux** variable (e.g., `nee`, `fc`, `gpp`) and optionally one meteorological variable (e.g., `ta`, `le`).

```{r choose_vars, echo=TRUE, eval=FALSE}
vars <- c(
  "fc",  # e.g., "fc"
  "ta",  # e.g., "ta"
  "gpp"   # optional third variable
)

# CHECK: confirm these columns exist (should return character(0))
setdiff(vars, names(df))
```


## 2) Compute population parameters (partial pipeline)

Complete the code to compute **population mean** and **population SD** for each selected variable.

Your output should be a tibble named `pop` with columns: `var`, `mean`, `sd`.

```{r pop_params, echo=TRUE, eval=FALSE}
# GOAL: Create a tidy table of population parameters for vars.
# TODO: complete this pipeline.

pop <- df %>%
  summarise(across(
    all_of(vars),
    list(
      mean = ~mean(.x, na.rm = TRUE),
      sd   = ~sd(.x, na.rm = TRUE)
    ),
    .names = "{.col}__{.fn}"
  )) %>%   
  pivot_longer(
    everything(),
    names_to = c("var", "stat"),
    names_sep = "__",
    values_to = "value"
  ) %>%   
  pivot_wider(names_from = "stat", values_from = "value")       # TODO: pivot_wider

pop
```

**Prompt (2–3 sentences):** Which variable has the largest population SD? What does that imply about the variability of that process?

> *Of the three variables that I chose, GPP has the highest sd. This implies that it has the highest variability, although TA is less 0.1 lesser. The lowest is FC, meaning that it likely has the lowest variability.*

---

# Relative error for point estimates (accuracy)

We will estimate the population mean with a sample mean, then compute **relative error**:

\[
\text{Relative error (\%)} = 100 \times \frac{|\mu - \bar{x}|}{|\mu|}
\]

## 3) One sample: compute relative error 

Write code to:

1. set a seed
2. choose a sample size `n_samp`
3. draw a simple random sample `samp`
4. compute `sample_mean` for each variable in `vars`
5. join to population means from `pop`
6. compute `rel_error_pct` for each variable

Return a tibble with columns: `var`, `sample_mean`, `pop_mean`, `rel_error_pct`.

```{r one_sample_relerr, echo=TRUE, eval=FALSE}
# GOAL: Relative error for point estimates from one random sample.

set.seed(100000)
n_samp <- 500

samp <- df %>%
  slice_sample(n = n_samp)

# get mean point estimate for each variable
est <- samp %>%
  summarise(across(all_of(vars), ~mean(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "var", values_to = "sample_mean")

# compute the relative error for each variable
relerr <- est %>%
  left_join(pop %>% select(var, pop_mean = mean), by = "var") %>%
  mutate(rel_error_pct = 100 * abs(pop_mean - sample_mean) / abs(pop_mean)) %>%
  arrange(desc(rel_error_pct))

relerr
```

**Prompt (3–4 sentences):** Which variable had the largest relative error at your chosen `n_samp`?  
Give a scientific reason why some variables are harder to estimate from a small sample.

> *Variable "fc" has the largest relative error by a large margin, reaching nearly 50%. Some variables may be hard to estimate from a small sample due to having a high variability relative to the sample size. When there is a lower population then it is also harder to distinguish patterns in the data, because we are using a relatively low amount of points. As a result the sub-sample from the population may not be totally representative of the whole. This can result in high error.*

---

## 4) Relative error vs sample size (partial code + you complete)

We will focus on **one variable** and simulate many repeated studies. The next code chunk is doing quite a lot. Please review the #comments throughout and check you understand generally what each section is doing.

```{r relerr_by_n, echo=TRUE, eval=FALSE}
set.seed(10000)

n_grid <- c(25, 100, 300, 1000)   # e.g., c(25, 100, 300, 1000)
reps <- 1000                        # e.g., 500 or 1000

target_var <- "fc"                # e.g., "fch4" or "fc"

# Pull the population vector (remove missing values once, up front)
x_pop <- df[[target_var]]
x_pop <- x_pop[!is.na(x_pop)]

# Population mean for the chosen variable
pop_mean_target <- pop %>%
  filter(var == target_var) %>%
  pull(mean)

# ---- Simulation using explicit for loops ----
sim_list <- vector("list", length(n_grid))

for (i in seq_along(n_grid)) {

  n_samp <- n_grid[i]

  # store relative error from each repeated "study"
  relerrs <- numeric(reps)

  for (j in seq_len(reps)) {

    x_samp <- sample(x_pop, size = n_samp, replace = FALSE)
    xbar <- mean(x_samp)

    relerrs[j] <- 100 * abs(pop_mean_target - xbar) / abs(pop_mean_target)
  }

  sim_list[[i]] <- tibble(
    n = n_samp,
    rel_error_pct = relerrs
  )
}

sim_relerr <- bind_rows(sim_list)

sim_relerr
```

### Summarize + plot (from scratch)

1) Summarize the simulated relative error by `n` using:
- median
- 90th percentile

2) Make a plot that communicates “relative error decreases with n”.

```{r summarize_plot_relerr, echo=TRUE, eval=FALSE}
# GOAL: Summarize and plot the relationship between n and relative error.
# TODO: write your own code.

summary_relerr <- sim_relerr %>% 
  group_by(n) %>% 
  summarise(
    median_relerr = median(rel_error_pct),
                           p90_relerr = quantile(rel_error_pct, 0.90)
  )
summary_relerr

p_relerr <- ggplot(summary_relerr, aes(x = n, y = median_relerr)) +
  geom_line(linewidth = 1.2, aes(color = "red")) +
  geom_point(aes(color = "red")) +
  geom_point(aes(y = p90_relerr, color = "blue")) +
  geom_line(aes(y = p90_relerr, color = "blue"), linetype = "dashed") +
  scale_x_log10() +
  theme_minimal() +
  labs(
    x = "Sample Size",
    y = "Relative Error (%)",
    title = "Relative Error vs. Sample Size"
  ) +
   scale_color_manual(values = c("red", "blue"), # Assign colors
                     labels = c('Median Error', '90th percentile error'))  # Assign new labels
p_relerr


summary_plot <- summary_relerr %>%
  pivot_longer(
    cols = c(median_relerr, p90_relerr),
    names_to = "stat",
    values_to = "rel_error_pct"
  )

p_relerr2 <- summary_plot %>%
  ggplot(aes(x = n, y = rel_error_pct, group = stat, color = stat)) + 
  geom_line(linewidth = 1.1) +
  geom_point() +
  theme_classic(base_size = 18) +
  labs(
    x = "Sample Size",
    y = "Relative Error %",
    title = ("Relative Error decreases with n")
  )

p_relerr2

```

**Prompt (3–4 sentences):** Describe the pattern. Is improvement from n=25→100 larger than 300→1000?  
Why does accuracy show “diminishing returns” as n increases?

> *The improvement from 25 to 100 is larger compared to the improvement of 300 to 1000. As you increase the sample size the error begins to decrease as well. However, we reach a point where the sample size is fairly representative of the population. Because of that there increasing the sample size no longer bears much improvement, as we see here.*

---

# Standard error and sampling distributions

For the sample mean, the standard error is approximately:

\[
SE(\bar{x}) \approx \frac{\sigma}{\sqrt{n}}
\]

where \(\sigma\) is population SD.

## 5) Theory vs simulation (partial pipeline)

Compute:
- theoretical SE
- simulated SD of sample means (over many replications)

```{r se_compare, echo=TRUE, eval=FALSE}
set.seed(480)

# TODO: pick a variable and sample size
target_var <- "gpp"
n_samp <- 300
reps <- 3000

pop_sd_target <- pop %>% filter(var == target_var) %>% pull(sd)
pop_mean_target <- pop %>% filter(var == target_var) %>% pull(mean)

means <- replicate(
  reps,
  mean(sample(df[[target_var]], size = n_samp, replace = FALSE), na.rm = TRUE)
)

se_theory <- pop_sd_target / sqrt(n_samp)
se_sim    <- sd(means)

tibble(
  target_var = target_var,
  n = n_samp,
  se_theory = se_theory,
  se_simulated = se_sim
)
```

### Plot the sampling distribution (fill-in)

```{r plot_sampling_dist, echo=TRUE, eval=FALSE}
tibble(mean_est = means) %>%
  ggplot(aes(x = mean_est)) +
  geom_histogram(bins = 50, alpha = 0.85) +
  geom_vline(xintercept = pop_mean_target, linetype = "dashed", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Sample mean of ", target_var),
    y = "Count",
    title = "Sampling distribution of the mean",
    subtitle = paste0("n = ", n_samp, "; reps = ", reps, " (dashed = population mean)")
  )
```

**Prompt (3–5 sentences):** How close were `se_theory` and `se_simulated`?  
What changes if you increase `n_samp` vs increase `reps`?

> *With my original setup up, n_samp = 300, reps = 3000, the se_theory and se_simulated are very close to one another. At first chaning the amount of reps did nothing, but by then reducing both the reps and n_samp, and changing one at a time I saw a pattern. Increasing the amount of reps decreases the difference between to the se_theory and se_simulated. Increasing the n_samp results in increasing the difference between the se_theory and se_simulated.*

---

# Bootstrap intuition (one sample, many resamples)

Bootstrap resampling approximates uncertainty **when you only have one sample**.

Key idea: treat your sample as a stand-in for the population and resample **with replacement**.

## 6) Bootstrap distribution of the mean 

Write code to:

- take ONE sample of size `n_samp` from `df`
- generate `B` bootstrap resamples (with replacement) from that sample
- compute bootstrap means
- compute bootstrap SD
- plot the bootstrap distribution

```{r bootstrap_mean, echo=TRUE, eval=FALSE}
set.seed(10000)

# TODO: choose settings
target_var <- "gpp"
n_samp <- 300
B <- 3000

# one "field campaign" sample
samp <- tibble(x = sample(x_pop, size = n_samp, replace = FALSE))

# bootstrap sampling of the mean
boot_means <- replicate(
  B,
  mean(sample(samp$x, size = n_samp, replace = TRUE))
)

boot_sd <- sd(boot_means)

tibble(
  sample_mean = mean(samp$x),
  boot_sd = boot_sd
)
```

```{r plot_boot, echo=TRUE, eval=FALSE}
# TODO: plot boot_means and mark the sample mean
tibble(boot_mean = boot_means) %>%
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 50, alpha = 0.85) +
  geom_vline(xintercept =  mean(samp$x), linetype = "dashed", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", target_var),
    y = "Count",
    title = "Bootstrap distribution of the mean",
    subtitle = paste0("n = ", n_samp, "; reps = ", reps, " (dashed = sample mean)")
  )
```

**Prompt (4–6 sentences):** Compare the bootstrap distribution to the sampling distribution from Part III.  
What is the bootstrap trying to approximate? When is bootstrap especially useful in EAES?

> *The bootstrap distribution is distributed similarly to the sampling distribution from part 3. However the sample mean and the bootstrap means are fairly different from one another from my results, however based on the intention of the bootstrap this is likely caused by an error in my code. The intention of the bootstrap distribution is to mimic the sampling distribution, and increase the sample size artificially. This can be especially useful in EAES in situations where our data is limited. In scenarios where there is limited time, funds, or another limiting factor then sampling may not be satisfactory and we may not be able to pick up on ths statistical trends from the samples alone. In these cases, bootstrapping can be applied*

---




# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` to your GitHub repo.
